Evaluating FAIR Digital Object as a distributed object system

### 2023-01-25 latex/table1.tex
Fixing references to use bibliography

\caption{Considering FDO and Web according to the quality levels of the Interoperability Framework for Fast Data (\protect\hyperlink{ref-KlCFFFyL}{Delgado, 2016}).

@@

\textbf{Symbiotic}: \emph{to what extent multiple applications can agree to interact/align/collaborate/cooperate} & The purpose of FDO is to enable federated machine actionable digital objects for scholarly purposes, in practice this also requires agreement of or compatibility between FDO types. FDO encourages research communities to develop common type registries to be shared across instances. In current DOIP practice, each service have their own types, attributes and operations. The wider symbiosis is consistent use of PIDs. & The Web is loosely coupled and encourages collaboration and linking by URL. In practice, REST APIs (\protect\hyperlink{ref-174AwcFUL}{Fielding, 2000}) end up being mandated centrally by dominant (often commercial) providers (\protect\hyperlink{ref-11VxBOeZy}{Fielding et al., 2017}), which clients are required to use as-is with special code per service. Use of Linked Data enables common tooling and semantic mapping across differences. \\
\textbf{Pragmatic}: \emph{using interaction contracts so processes can be choreographed in workflows} & FDO types and operations enable detailed choreography (see CWFP). \texttt{0.TYPE/DOIPOperation} has lightweight definition of operation, \texttt{0.DOIP/Request} or \texttt{0.DOIP/Response} may give FDO Type or any other kind of ``specifics'' (incl.~human readable docs). Semantics/purpose of operations not formalised (similar operations can be grouped with \texttt{0.DOIP/OperationReference}). & ``Follow your nose'' crawler navigation, which may lead to frequent dead ends. Operational composition, typically within a single API provider, documented by OpenAPI 3 (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}), schema.org Actions (\protect\hyperlink{ref-dKAekUjL}{{`Schema.org Actions - schema.org'}}), WSDL/SOAP (\protect\hyperlink{ref-LmudecyN}{{`Web Services Description Language (WSDL) Version 2.0 Part 0: Primer'}}), but frequently just as human-readable developer documentation/examples. \\
\textbf{Semantic}: \emph{ensuring consistent understanding of messages, interoperability of rules, knowledge and ontologies} & FDO semantic enable navigation and typing. Every FDO has a type. Types maintained in FDO Type registries, which may add additional semantics, e.g.~the ePIC \href{https://hdl.handle.net/21.11104/c1a0ec5ad347427f25d6}{PID-InfoType for Model}. No single type semantic, Type FDOs can link to existing vocabularies \& ontologies. JSON-LD used within some FDO objects (e.g.~DISSCO Digital Specimen, NIST Material Science schema) (\protect\hyperlink{ref-jRWurmQc}{Wittenburg et al., 2022}) & Lightweight HTTP semantics for authenticity/navigation. Semantic Type not commonly expressed on PID/header level, may be declared within Linked Data metadata. Semantic of type implied by Linked Data formats (e.g.~OWL2, RDFS), although choice of type system may not be explicit. \\

@@

\textbf{Connective}: \emph{transferring messages to another application, e.g.~wrapping in other protocols} & DOIP (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) is transport-independent, commonly TLS TCP/IP port 9000), \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP} & HTTP/1.1 (TCP/IP port 80), HTTP/1.1+TLS (TCP/IP 443), HTTP/2 (as HTTP/1* but binary), HTTP/3 (like HTTP/2+TLS but UDP) \\

>>>>

\caption{Considering FDO and Web according to the quality levels of the Interoperability Framework for Fast Data \cite{KlCFFFyL}.

@@

\textbf{Symbiotic}: \emph{to what extent multiple applications can agree to interact/align/collaborate/cooperate} & The purpose of FDO is to enable federated machine actionable digital objects for scholarly purposes, in practice this also requires agreement of or compatibility between FDO types. FDO encourages research communities to develop common type registries to be shared across instances. In current DOIP practice, each service have their own types, attributes and operations. The wider symbiosis is consistent use of PIDs. & The Web is loosely coupled and encourages collaboration and linking by URL. In practice, REST APIs \cite{174AwcFUL} end up being mandated centrally by dominant (often commercial) providers \cite{11VxBOeZy}, which clients are required to use as-is with special code per service. Use of Linked Data enables common tooling and semantic mapping across differences. \\
\textbf{Pragmatic}: \emph{using interaction contracts so processes can be choreographed in workflows} & FDO types and operations enable detailed choreography (see CWFP). \texttt{0.TYPE/DOIPOperation} has lightweight definition of operation, \texttt{0.DOIP/Request} or \texttt{0.DOIP/Response} may give FDO Type or any other kind of ``specifics'' (incl.~human readable docs). Semantics/purpose of operations not formalised (similar operations can be grouped with \texttt{0.DOIP/OperationReference}). & ``Follow your nose'' crawler navigation, which may lead to frequent dead ends. Operational composition, typically within a single API provider, documented by OpenAPI 3 \cite{k0AfCGzw}, schema.org Actions \cite{dKAekUjL}, WSDL/SOAP \cite{LmudecyN}, but frequently just as human-readable developer documentation/examples. \\
\textbf{Semantic}: \emph{ensuring consistent understanding of messages, interoperability of rules, knowledge and ontologies} & FDO semantic enable navigation and typing. Every FDO has a type. Types maintained in FDO Type registries, which may add additional semantics, e.g.~the ePIC \href{https://hdl.handle.net/21.11104/c1a0ec5ad347427f25d6}{PID-InfoType for Model}. No single type semantic, Type FDOs can link to existing vocabularies \& ontologies. JSON-LD used within some FDO objects (e.g.~DISSCO Digital Specimen, NIST Material Science schema) \cite{jRWurmQc} & Lightweight HTTP semantics for authenticity/navigation. Semantic Type not commonly expressed on PID/header level, may be declared within Linked Data metadata. Semantic of type implied by Linked Data formats (e.g.~OWL2, RDFS), although choice of type system may not be explicit. \\

@@

\textbf{Connective}: \emph{transferring messages to another application, e.g.~wrapping in other protocols} & DOIP \cite{13TcbsZF6} is transport-independent, commonly TLS TCP/IP port 9000), \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP} & HTTP/1.1 (TCP/IP port 80), HTTP/1.1+TLS (TCP/IP 443), HTTP/2 (as HTTP/1* but binary), HTTP/3 (like HTTP/2+TLS but UDP) \\

### 2023-01-25 latex/table3.tex
Fixing references to use bibliography

  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1642}}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
\caption{Comparing FAIR Digital Object (with the DOIP 2.0 protocol (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018})) and Web technologies (using Linked Data) as middleware infrastructures (\protect\hyperlink{ref-hRzcHhPD}{Zarras, 2004})
\label{tbl:fdo-web-middleware}}\tabularnewline

@@

\begin{minipage}[b]{\linewidth}\raggedright
\emph{Quality}

@@

FDO w/ DOIP

@@

Web w/ Linked Data

@@

\textbf{Openness}: \emph{framework enable extension of applications} & FDOs can be cross-linked using PIDs, pointing to multiple FDO endpoints. Custom DOIP operations can be exposed, although it is unclear if these can be outside the FDO server. PID minting requires Handle.net prefix subscription, or use of services like \href{https://datacite.org/}{Datacite}, \href{https://eudat.eu/services/userdoc/b2handle}{B2Handle}. & The Web is inherently open and made by cross-linked URLs. Participation requires DNS domain purchase (many free alternatives also exists). PID minting can be free using PURL/ARK services, or can use DOI/Handle with HTTP redirects. \\
\textbf{Scalability}: \emph{application should be effective at many different scales} & No defined methods for caching or mirroring, although this could be handled by backend, depending on exposed FDO operations (e.g.~Cordra can scale to multiple backend nodes) & Cache control headers reduce repeated transfer and assist explicit and transparent proxies for speed-up. HTTP \texttt{GET} can be scaled to world-population-wide with Content-Delivery Networks (CDNs), while write-access scalability is typically manage by backend. \\
\textbf{Performance}: \emph{efficient and predictable execution} & DOIP has been shown moderately scalable to 100 millions of objects, create operation at 900 requests/second . DOIP protocol is reusable for many operations, multiple requests may be answered out of order (by \texttt{requestId}). Multiple connections possible. Setup is typically through TCP and TLS which adds latency. & HTTP traffic is about 10\% of global Internet traffic, excluding video and social networks (\protect\hyperlink{ref-F8VZ86hu}{Sandvine}). HTTP 1 connections are serial and reusable, and concurrent connections is common. HTTP/2 adds asynchronous responses and multiplexed streams (\protect\hyperlink{ref-KbbW0kGT}{Belshe \& Peon, 2015}) but still has TCP+TLS startup costs. For reduced latency , HTTP/3 (\protect\hyperlink{ref-Amit86Tp}{{`draft-ietf-quic-http-34'}}) use QUIC (\protect\hyperlink{ref-1GWvyEmUf}{Iyengar \& Thomson, 2021})) rather than TCP, already adapted heavily (30\% of EMEA traffic) of which \href{https://engineering.fb.com/2020/10/21/networking-traffic/how-facebook-is-bringing-quic-to-billions/}{Instagram \& Facebook video} is the majority of traffic. \\
\textbf{Distribution transparency}: \emph{application perceived as a consistent whole rather than independent elements.} & Each FDO is accessed separately along with its components (typically from the same endpoint). FDOs should provide the mandatory kernel metadata fields. FDOs of the same declared type typically share additional attributes (although that schema may not be declared). DOIP does not enforce metadata typing constraints, this need to be established as FDO conventions. & Each URL accessed separately. Common HTTP headers provide basic metadata, although it is often not reliable. A multitude of schemas and serializations for metadata exists, conventions might be implied by a declared profile or certain media types. Metadata is not always machine findable, may need pre-agreed API URI Templates (\protect\hyperlink{ref-11hxwwuRt}{Gregorio et al., 2012}), content-negotiation (\protect\hyperlink{ref-n2f3TfoI}{{`Content negotiation - HTTP \textbar{} MDN'}}) or FAIR Signposting (\protect\hyperlink{ref-snykkm7R}{Van de Sompel et al., 2022}). \\
\textbf{Access transparency}: \emph{local/remote elements accessed similarly} & FDOs should be accessed through PID indirection, this means difficult to make private test setup. Commonly a fixed DOIP server is used directly, which permits local non-PID identifiers. & Global HTTP protocol frequently used locally and behind firewalls, but at risk of non-global URIs (e.g.~\texttt{http://localhost/object/1}) and SSL issues (e.g.~self-signed certificates, local CAs) \\
\textbf{Location transparency}: \emph{elements accessed without knowledge of physical location} & FDOs always accessed through PIDs. Multiple locations possible in Handle system, can expose geo-info. & PIDs and URL redirects. DNS aliases and IP routing can hide location. Geo-localised servers common for large cloud deployments. \\
\textbf{Concurrency transparency}: \emph{concurrent processing without interference} & No explicit concurrency measures. FDO kernel metadata can include checksum and date. & HTTP operations are classified as being stateless/idempotent or not (e.g.~\texttt{PUT} changes state, but can be repeated on failure), although these constraints are occassionally violated by Web applications. Cache control, \texttt{ETag} (\textasciitilde{} checksum) and modification date in HTTP headers allows detection of concurrent changes on a single resource. \\
\textbf{Failure transparency}: \emph{service provisioning resilient to failures} & DOIP status codes, e.g.~\texttt{0.DOIP/Status.104}, additional codes can be added as custom attributes & HTTP \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-6.5}{status codes} e.g.~\texttt{404\ Not\ Found}, specific meaning of standard codes can be \href{https://swagger.io/docs/specification/describing-responses/}{documented in Open API}. Custom codes uncommon. \\
\textbf{Migration transparency}: \emph{allow relocating elements without interfering application} & Update of PID record URLs, indirection through \texttt{0.TYPE/DOIPServiceInfo} (not always used consistently). No redirection from DOIP service. & HTTP \texttt{30x} status codes provide temporary or permanent redirections, commonly used for PURLs but also by endpoints. \\
\textbf{Persistence transparency}: \emph{conceal deactivation/reactivation of elements from their users} & FDO requires use of PIDs for object persistence, including a tombstone response for deleted objects. There is no guarantee that an FDO is immutable or will even stay the same type (note: CORDRA extends DOIP with \href{https://www.cordra.org/documentation/design/object-versioning.html}{version tracking}). & URLs are not required to persist, although encouraged (\protect\hyperlink{ref-rbG9uRKw}{{`Hypertext Style: Cool URIs don't change.'}}). Persistence requires convention to use PIDs/PURLs and HTTP \texttt{410\ Gone}. An URL may change its content, change in type may sometimes force new URLs if exposing extensions like \texttt{.json}. Memento (\protect\hyperlink{ref-DNjcMi4a}{Van de Sompel, Nelson \& Sanderson, 2013}) expose versioned snapshots. WebDAV \texttt{VERSION-CONTROL} method (\protect\hyperlink{ref-hRUsaTiz}{Clemm et al., 2002}) (used by SVN). \\
\textbf{Transaction transparency}: \emph{coordinate execution of atomic/isolated transactions} & No transaction capabilities declared by FDO or DOIP. Internal synchronisation possible in backend for Extended operations. & Limited transaction capabilities (e.g.~\texttt{If-Unmodified-Since}) on same resource. WebDAV \href{https://datatracker.ietf.org/doc/html/rfc4918\#section-6}{locking mechanisms} (\protect\hyperlink{ref-xvMAyRAc}{Dusseault, 2007}) with \texttt{LOCK} and \texttt{UNLOCK} methods. \\
\textbf{Modularity}: \emph{application as collection of connected/distributed elements} & FDOs are inheritedly modular using global PID spaces and their cross-references. In practice, FDOs of a given type are exposed through a single server shared within a particular community/institution. & The Web is inheritently modular in that distributed objects are cross-referenced within a global URI space. In practice, an API's set of resources will be exposed through a single HTTP service, but modularity enables fine-grained scalability in backend. \\
\textbf{Encapsulation}: \emph{separate interface from implementation. Specify interface as contract, multiple implementations possible} & Indirection by PID gives separation. FDO principles are protocol independent, although it may be unclear which protocol to use for which FDO (although \texttt{0.DOIP/Transport} can be specified after already contacting DOIP). Cordra supports \href{https://www.cordra.org/documentation/api/doip.html}{native DOIP}, \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP} and \href{https://www.cordra.org/documentation/api/rest-api.html}{Cordra REST API}) & HTTP/1.1 semantics can seemlessly upgrade to HTTP/2 and HTTP/3. \texttt{http} vs \texttt{https} URIs exposes encryption detail\footnote{The \texttt{http} protocol (port 80) can in theory also upgrade (\protect\hyperlink{ref-1B16WfR7q}{Khare \& Lawrence, 2000}) to TLS encryption, as commonly used by \href{https://www.rfc-editor.org/rfc/rfc8010.html\#section-8.2}{Internet Printing Protocol} for \texttt{ipp} URIs, but on the Web, best practice is explicit \texttt{https} (port 443) URLs to ensure following links stay secure.}. Implementation details may leak into URIs (e.g.~\texttt{search.aspx}), countered by deliberate design of URI patterns (\protect\hyperlink{ref-iME5l1kk}{{`Hypertext Style: Cool URIs don't change.'}}) and PIDs via Persistent URLs (PURL). \\
\textbf{Inheritance}: \emph{Deriving specialised interface from another type} & DOIP types nested with parents, implying shared FDO structures (unclear if operations are inherited). FDO establishes need for multiple Data Type Registries (e.g.~managed by a community for a particular domain). Semantics of type system currently undefined for FDO and DOIP, syntactic types can also piggyback of FDO type's schema (e.g.~\href{(https://www.cordra.org/documentation/design/schemas.html\#schema-references)}{CORDRA \texttt{\$ref}} use of \href{https://json-schema.org/draft/2020-12/json-schema-core.html\#references}{JSON Schema references} (\protect\hyperlink{ref-15EZ2D0Rm}{{`draft-bhutton-json-schema-00'}})) & Syntactically Media Type with multiple suffixes (\protect\hyperlink{ref-nlH2rm1s}{{`draft-ietf-mediaman-suffixes-00 - Media Types with Multiple Suffixes'}}) (mainly used with \texttt{+json}), declaration of subtypes as profiles (RFC6906) (\protect\hyperlink{ref-nCzRO0pK}{Wilde, 2013}). In metadata, semantic type systems (RDFS (\protect\hyperlink{ref-ZwAcGQKY}{{`RDF Schema 1.1'}}), OWL2 (\protect\hyperlink{ref-1p4IWJpI}{{`OWL 2 Web Ontology Language Document Overview (Second Edition)'}}), SKOS (\protect\hyperlink{ref-15gQDya5B}{{`SKOS Simple Knowledge Organization System Primer'}})). OpenAPI 3 (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}) \href{https://spec.openapis.org/oas/v3.1.0\#composition-and-inheritance-polymorphism}{inheritance and Polymorphism}. XML \texttt{xsd:schemaLocation} or \texttt{xsd:type} (\protect\hyperlink{ref-11oVTMsuZ}{{`W3C XML Schema Definition Language (XSD) 1.1 Part 1: Structures'}}), JSON \texttt{\$schema} (\protect\hyperlink{ref-15EZ2D0Rm}{{`draft-bhutton-json-schema-00'}})), JSON-LD \texttt{@context} (\protect\hyperlink{ref-by6CfOmv}{{`JSON-LD 1.1'}}). Large number of domain-specific and general ontologies define semantic types, but finding and selecting remains a challenge. \\
\textbf{Signal interfaces}: \emph{asynchronous handling of messages} & DOIP 2.0 is synchronous, in FDO async operations undefined. Could be handled as custom jobs/futures FDOs & HTTP/2 \href{https://datatracker.ietf.org/doc/html/rfc7540\#section-5}{multiplexed streams} (\protect\hyperlink{ref-KbbW0kGT}{Belshe \& Peon, 2015}), Web Sockets (\protect\hyperlink{ref-qgqADk45}{{`WebSockets Standard'}}), Linked Data Notifications (\protect\hyperlink{ref-zlGYiuWC}{{`Linked Data Notifications'}}), AtomPub (\protect\hyperlink{ref-IbaebabD}{Gregorio \& de hOra, 2007}), SWORD (\protect\hyperlink{ref-7DGllKzG}{{`SWORD 3.0 Specification'}}), Micropub, more typically ad-hoc jobs/futures REST resources \\
\textbf{Operation interfaces}: \emph{defining operations possible on an instance, interface of request/response messages} & CRUD predefined in DOIP, custom operations through \texttt{0.DOIP/Op.ListOperations} (can be FDOs of type \texttt{0.TYPE/DOIPOperation}, more typically local identifiers like \texttt{"getProvenance"}) & CRUD predefined in \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-4.3}{HTTP methods} (\protect\hyperlink{ref-HE7Ikwwl}{Fielding \& Reschke, 2014a}), (\href{https://www.iana.org/assignments/http-methods/http-methods.xhtml}{extended by registration}), URI Templates (\protect\hyperlink{ref-11hxwwuRt}{Gregorio et al., 2012}), \href{https://spec.openapis.org/oas/v3.1.0.html\#operation-object}{OpenAPI operations} (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}), HATEOAS\footnote{HATEOAS: Hypermedia as the Engine of Application State (\protect\hyperlink{ref-174AwcFUL}{Fielding, 2000}), an important element of the REST architectural style.} incl.~Hydra (\protect\hyperlink{ref-13PaKBrZe}{{`Hydra W3C Community Group'}}), schema.org Actions {[}{`Schema.org Actions - schema.org'}), JSON HAL (\protect\hyperlink{ref-nyrPYHoP}{{`draft-kelly-json-hal-08'}}) \& Link headers (RFC8288) (\protect\hyperlink{ref-ozdWB3O7}{Nottingham, 2017}) \\
\textbf{Stream interfaces}: \emph{operations that can handle continuous information streams} & Undefined in FDO. DOIP can support multiple byte stream elements (need custom FDO type to determine stream semantics) & HTTP 1.1 (\protect\hyperlink{ref-tIkSXrb5}{Fielding \& Reschke, 2014b}) \href{https://datatracker.ietf.org/doc/html/rfc7230\#section-4.1}{chunked transfer}, HLS (RFC8216) (\protect\hyperlink{ref-PL512B0X}{May, 2017}), MPEG-DASH (ISO/IEC 23009-1:2019) (\protect\hyperlink{ref-ikO7e0mh}{14:00-17:00}) \\

>>>>

  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3676}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1544}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1544}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1618}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1618}}@{}}
\caption{Checking FDO guidelines \cite{RwvirqWg, yygVPoL0} against its current implementations as DOIP \cite{13TcbsZF6} and Linked Data Platform (LDP) \cite{7szz7dwO}, with suggestions for required additions.
\label{tbl:fdo-checks}}\tabularnewline

@@

\begin{minipage}[b]{\linewidth}\centering
\textbf{FDO Guideline}

@@

DOIP 2.0

@@

FDO suggestions
Linked Data Platform
LDP suggestion

@@

G1: \emph{invest for many decades} & Handle system stable for 20 years, DOIP 2.0 since 2017. & Ensure FDO types will not be protocol-bound as DOIP might be updated/replaced & HTTP stable for 30 years, Semantic Web for 20 years. \texttt{http://} URIs replaced by \texttt{https://}. & Keep flexibility of RDF serialisation formats which may change more frequently \\
G2: \emph{trustworthiness} & DOI/Handle trusted by all major academic publishers and data repositories. DOIP relatively unknown, in effect only one implementation. & Further promote DOIP and justify its benefits. Build tutorials and OSI open source implementations. Standardise DOIP-over-HTTP alternative. & JSON-LD used by half of all websites \cite{j3SqQLBR}, however previous bad experiences with Semantic Web may deter adapters & Ensure simplicity for end developers, rather than semantic overengineering. Example-driven documentation. \\
G3: \emph{follows FAIR principles} & See Table \ref{tbl:fair-data-maturity-model} & Ensure all FAIR principles are covered, build complete examples. & Touched briefly, see Table \ref{tbl:fair-data-maturity-model} & Add explicit expression to show each FAIR pcinciple fulfilled. \\
G4: \emph{machine actionability} & CRUD and extension operations dynamically listed (see Table \ref{tbl:fdo-web-middleware}) & Specify which operations should work for a given type, to reduce need for dynamic lookup. Specify input/output expectations formally (e.g.~JSON Schema). & HTTP CRUD operations, Open API (see Table \ref{tbl:fdo-web-middleware}) & Document operations so client can make subsequent HTTP calls. \\
G5: \emph{abstraction principle} & Handle PIDs as abstraction base. DOIP operations can use any transport protocol. & Document transport protocols as FDOs, recommend which transport to use. & URI as abstraction base. Does not specify PID requirements. & Give stronger deployment recommendations. \\
G6: \emph{stable binding between entities} & Machine-navigation through PIDs and operations specified per type. Unclear when metadata field is a PID or plain text. & Make datatype of fields explicit to support navigation. & Machine-navigation through URIs via properties and types. Unclear when URI should be followed or is just identifier, but always distinct from plain text. & \\
G7: \emph{encapsulation} & Operations discovered at runtime (\texttt{0.DOIP/Op.ListOperations}). & Allow method discovery by type FDOs in advance (see PR-TypingFDOs-2.0-20220608). & HTTP methods discovered at runtime (\texttt{OPTIONS}), indempotent methods attempted directly. Unsupported methods reported using LDP constraints to human-readable text. & Declare supported methods in advance, e.g.~OpenAPI \cite{k0AfCGzw} \\
G8: \emph{technology independence} & In theory independent, in reality depends on single implementations of Handle system and DOIP & Encourage open source DOIP testbeds and lighter reference implementations & Multiple HTTP implementations, multiple LDP implementations. No FDOF implementations. & Develop demonstrator of FDOF usage based on existing LDP server. \\
G9: \emph{standard compliance} & Handle \cite{2gXY2Nwh}, DOIP \cite{13TcbsZF6}. FDO requirements not standardised yet. & Formalise standard process of FDO requirements \cite{TQdku4YF} & HTTP, LDP. FDOF not yet standardised & Formalise FDOF from FDOF-SEM working group \\
FDOF1: \emph{PID as basis} & Extensive use of Handle system. & Clarify how local testing handles can be used during development, how ``temporary'' FDOs should evolve \cite{ljVViWCl}. Register \texttt{0.DOIP/*} and \texttt{0.FDO/*} as PIDs. & HTTP URLs as basis for identifiers, but they are frequently not persistent. & Add strong guidance for PID services like w3id and persistence policies. \\
FDOF2: \emph{PID record w/ type} & Unspecified how to resolve from Handle to DOIP Service (!), in practice \texttt{10320/loc}, \texttt{0.TYPE/DOIPService}, \texttt{URL}, \texttt{URL\_REPLICA} & Document requirements for PID Record & w3id/purl PIDs redirect without giving any metadata. Datacite DOIs content-negotiate to give registered metadata. & Add FAIR Signposting at PID provider for minimal PID record \\
FDOF3: \emph{PID resolvable to bytestream \& metadata} & Byte stream resolvable (\texttt{0.DOIP/Retrieve}), \texttt{includeElementData} option can retrieve bytestream or full object structure. No method/attribute defined for separate metadata, only directly in PID Record. Unclear meaning of multiple items and bytestream chunks. & Clarify expectations for multiple items. Recommend chunks to not be used. & URIs resolvable by default. Multiple ways to resolve metadata, unclear preference. & Add FAIR Signposting and preference order. \\
FDOF4: \emph{Additional attributes} & Freetext attribute keys. Attributes should be defined for FDO type (?). & Require that attribute keys should be PIDs (or have predefined mapping to PIDs). Explicitly allow attributes not already defined in type. & All attributes individually identified. Any Linked Data attributes can be used by URI or with mapped prefix. & Clarify type expectations of required/recommended/optional attributes. \\
FDOF5: \emph{Interface: operation by PID} & Extended operations use PID, but ``pid-like'' DOIP operations/types are not registered as handles. & Register \texttt{0.DOIP/*} and \texttt{0.FDO/*} as PIDs. Clarify that operations can be mapped to protocol directly. & CRUD operations used directly in HTTP (e.g.~\texttt{PUT}). Unclear how to provide PID for additional operations. & Specify how additional operations should be called over HTTP. \\
FDOF6: \emph{CRUD operations + extensions} & \texttt{0.DOIP/Op.Create}, \texttt{Op.Retrieve}, \texttt{Op.Update}, \texttt{Op.Delete} but also \texttt{0.DOIP/Op.Search}. & Document & \texttt{PUT}, \texttt{GET}, \texttt{POST}, \texttt{DELETE}, \texttt{PATCH}, \texttt{HEAD} -- extension operations (e.g.~WebDAV \texttt{COPY}) not used, resource patterns \cite{i9Ang0rM} are used instead. & Document how operation resources can be discovered from an LPD container. Document search API. \\
FDOF7: \emph{FDOF Types related to operations} & Not yet formalised, by DOIP discoverable on a given FDO rather than type. PR-TypingFDOs leaves this open. & Add explicit relation between type and operations & \texttt{OPTIONS} per LDP Resource, but not by type. Common types (\texttt{ldp:Resource}, \texttt{ldp:Container}) indicate LDP support, but are not required. & Always make LDP types explicit in FDO profile. \\
FDOF8: \emph{Metadata as FDO, semantic assertions} & DOIP includes all metadata in PID Record. Separate Metadata FDO need custom property. & Specify a \texttt{0.FDO/metadata} or similar to point to Metadata FDOs. & Assertions are always with semantics, using RDF vocabularies. Unspecified how to find additional metadata resources, \texttt{rdfs:seeAlso} is common. & Use FAIR Signposting \texttt{describedby} link relation to additional metadata PIDs \\
FDOF9: \emph{Different metadata levels} & Defines open-ended ``Response Attributes'' without namespaces, but mandated as ``None'' for all CRUD operations. Metadata would need to be bundled within custom FDO types/attributes. Unclear how levels are separated within single FDO representation (need FDOF8?). & Declare which metadata are expected within response attribute or within FDO object. Require PIDs for custom attributes. Define how alternate metadata levels can be represented separately. & Undefined how to handle multiple metadata granularities or domains, alternative LDP containers can present different views on same stored objects. & Define how to navigate to alternate views and their semantic implications, e.g.~\texttt{owl:sameAs} \\
FDOF10: \emph{Metadata schemas by community} & Metadata schemas are in practice managed on single CORDA server as local types, using JSON Schema. & Require types to be FDOs with registered PIDs, implement shared types. & Plethora of existing RDF vocabularies/ontologies managed by larger communities, e.g.~\href{https://obofoundry.org/}{OBO Foundry} \cite{CNP39tah} & Rather document better how individual ad-hoc schemas can be started for prototypes. \\
FDOF11: \emph{FDO collections w/ semantic relations} & Collection type undefined by DOIP. Informal use of \texttt{HAS\_PARTS} Handle attribute (e.g. \cite{6LvydTsC}). & & LDP Containers required by specification, also user-created (eg. \texttt{BasicContainer}). & Clarify relation to other collections like DCAT 3 \cite{1EW4hLN8U}, \href{https://schema.org/Dataset}{Schema.org Dataset}, OAI-ORE \cite{gtGuSevK} \\
FDOF12: \emph{Deleted FDO preserve PID w/ tombstone} & Tombstone for deleted resource undefined by DOIP. \texttt{0.DOIP/Status.104} status code does not distinguish ``Not Found'' or ``Gone'' & Formalise tombstone requirements with new FDO type & \texttt{410\ Gone} recommended, but \texttt{404\ Not\ Found} common. No requirement for tombstone serialisation & Formalise tombstone requirements and serialisation \\
